---
title: "sdhmR Homework 4_M2.4_Creating Pseudo-Absences in R"
author: "Eli L. Polzer"
date: "18 February 2021"
output: html_document
---

<!-- 

-->
```{r global_options, include=FALSE}

```

---

## Context:
The data are of the *Pinus edulis* Common piñon, the species you dealt with in Exercise #3.  In this exercise we assume the acquisition of species location data reveals that data are presence-only (measured 1's only; no 0=absence)

The goal is to build a dataframe consisting of pseudo--absences selected from *edulis*' geographic and environmental background.

---

## The Data:
* Common piñon *Pinus edulis* sppcode=106
* Data for each group in:
  * **`~\sdhmR-V2020.2\data\exercise\traindat`** 
* Data file:
  * **`spp106pr_RANG.csv`**

---

## Question #1:
* Import the data type you have been assigned to work with
* Build a buffered-presence points bounding box
  * Buffer each point by ~50km 
  * Ensure you assign correct CRS, values, and resolution to bounding box
```{r}
path.root <- "~/sdhmR-V2021.1EXP"
path.dat <- paste(path.root, "C:/Users/epolzer/Documents/sdhmR-V2021.1/sdhmR-V2021.1EXP/Data/exercise/traindat", sep = "")
path.mod2 <- paste(path.root, "C:/Users/epolzer/Documents/sdhmR-V2021.1/sdhmR-V2021.1EXP/Data/module02", sep = "")
path.gis <- paste(path.root, "C:/Users/epolzer/Documents/sdhmR-V2021.1/sdhmR-V2021.1EXP/Data/gis_layers", sep = "")
path.figs <- paste(path.root, "C:/Users/epolzer/Documents/sdhmR-V2021.1/sdhmR-V2021.1EXP/Figures", sep = "")

setwd("~/sdhmR-V2021.1EXP")
source("C:/Users/epolzer/Documents/sdhmR-V2021.1/sdhmR-V2021.1EXP/R Code/prjs4_sdhmR.R")
ls(pattern = "prj.")

library(raster)
library(sf)

setwd("C:/Users/epolzer/Documents/sdhmR-V2021.1/sdhmR-V2021.1EXP/Data/exercise/traindat")
tru.range <- read.csv("spp106pr_RANG_2.csv")
dim(tru.range)
head(tru.range, 2)

pied.range <- st_as_sf(tru.range, coords = c("wgs_xF", "wgs_yF"), 
                      crs = prj.wgs84, remove = F) 
head(pied.range, 2)
class(pied.range)

extb <- extent(min(tru.range$wgs_xF) - 0.5, max(tru.range$wgs_xF) + 0.5, 
               min(tru.range$wgs_yF) - 0.5, max(tru.range$wgs_yF) + 0.5)
pied.range.bufSF <- st_as_sfc(st_bbox(extb, crs = prj.wgs84))
pied.range.bufSF

setwd("C:/Users/epolzer/Documents/sdhmR-V2021.1/sdhmR-V2021.1EXP/Data/gis_layers")
t1 <- raster("templateR_wgs.img")
t1

pied.range.bufSP <- as_Spatial(pied.range.bufSF) 
pied.range.bufR <- crop(rasterize(pied.range.bufSP, t1), pied.range.bufSP)
pied.range.bufR
```

---

## Question #2:
* Build a fishnet for the bounding box
```{r}
f1 <- sp::coordinates(pied.range.bufR) 
f2 <- cellFromXY(pied.range.bufR, f1) 
f3 <- as.data.frame(cbind(f1, f2)) 
head(f3, 2) 
tail(f3, 2) 
ncell(pied.range.bufR)

names(f3)[1:3] <- c("cell.wgs_x", "cell.wgs_y", "FNETID")
f3 <- f3[c("FNETID", "cell.wgs_x", "cell.wgs_y")] 

pied.range.fnetSF <- st_as_sf(f3, coords = c("cell.wgs_x", "cell.wgs_y"), 
                           crs = prj.wgs84, remove = F) 
head(pied.range.fnetSF, 2)

class(pied.range.fnetSF) 
pied.range.fnetDF <- st_drop_geometry(pied.range.fnetSF)
head(pied.range.fnetDF, 2)
tail(pied.range.fnetDF, 2)
class(pied.range.fnetDF)
```
---

## Question #3:
* Use the fishnet to create a pseudo--absence data frame of [X,Y]’s
* For simplicity’s sake, set $n$ for pseudo--absence at twice (2$\times$) the number of presences in your respective RANG, PERS, MORT, and SEED data set
* Bind these data to the presence dataframe you imported in Question #1
```{r}
setwd("C:/Users/epolzer/Documents/sdhmR-V2021.1/sdhmR-V2021.1EXP/Data/module02")
load("pied.framesR.RData")
pied.range.bufR

setwd("C:/Users/epolzer/Documents/sdhmR-V2021.1/sdhmR-V2021.1EXP/Data/module02")
load("pied.framesSF.RData")
pied.range.bufSF

load("pied.range.fnet.RData")
head(pied.range.fnetDF , 2)
head(pied.range.fnetSF , 2)

tru.xy <- tru.range[c("wgs_xF", "wgs_yF")]
pX <- sp::coordinates(tru.xy) 
pres.fnetid <- cellFromXY(pied.range.bufR, pX)
tru.rangeFNET <- cbind(pres.fnetid, tru.range)
names(tru.rangeFNET)[1] <- "FNETID"
head(tru.rangeFNET, 2)

buf.fnetid <- extract(pied.range.bufR, pied.range.fnetDF[c("cell.wgs_x", "cell.wgs_y")])
tru.bufFNET <- cbind(pied.range.fnetDF, buf.fnetid) 
head(tru.bufFNET, 2)

m1 <- merge(pied.range.fnetDF, tru.bufFNET, by = c("FNETID", "cell.wgs_x", "cell.wgs_y"), all.y = T)
head(m1, 2)
pied.indexFNET <- merge(m1, tru.rangeFNET, by = c("FNETID"), all = T) 
head(pied.indexFNET, 2)
names(pied.indexFNET)[4] <- "in.modFR"

setwd("C:/Users/epolzer/Documents/sdhmR-V2021.1/sdhmR-V2021.1EXP/Data/module02")
p1 <- get(load("pied.indexFNET.RData"))
head(p1, 2)

p2.spp <- subset(p1$FNETID, p1$RANG106 == 1)
p2.modFR <- subset(p1$FNETID, p1$in.modFR == 1)
length(p2.spp)
length(p2.modFR)

p3 <- p1[!p1$FNETID %in% p2.spp, ]
p4 <- p1[!p1$FNETID %in% p2.spp & p1$FNETID %in% p2.modFR, ]
psu.buf <- p4

set.seed(1234) 
psu.srs2 <- psu.buf[sample(1:nrow(psu.buf), 2 * table(p1$RANG106)[[1]], replace = F), ]
psu.srs2$RANG106 <- 0  
psu.srs2$in.modFR <- 1 
dim(psu.srs2) 
head(psu.srs2, 2)

head(tru.rangeFNET, 2)
head(psu.srs2, 2)
pied.PPsA <- merge(tru.rangeFNET, psu.srs2, by = c("FNETID", "RANG106", "wgs_xF", "wgs_yF"), all = T)
pied.PPsA$in.modFR <- NULL

pied.PPsA$tr.wgs_x <- ifelse(pied.PPsA$RANG106 == 0, pied.PPsA$cell.wgs_x, pied.PPsA$wgs_xF)
pied.PPsA$tr.wgs_y <- ifelse(pied.PPsA$RANG106 == 0, pied.PPsA$cell.wgs_y, pied.PPsA$wgs_yF)
```
---

## Question #4:
* Save your data as R objects, including both the true presence and pseudo-absences:
  * Dataframe;
  * Point shapefile with geometry in R
  * Export as a point shapefile in ESRI format
* Save these R objects in a **`.RData`** file
* Save the bounding box as a raster--based **`.img`** file
```{r}
setwd("C:/Users/epolzer/Documents/sdhmR-V2021.1/sdhmR-V2021.1EXP/Data/module02")
save("pied.range.bufR", file = "pied.framesR.RData")
save("pied.range.bufSF", file = "pied.framesSF.RData")
save("pied.range.bufSP", file = "pied.framesSP.RData")

setwd("C:/Users/epolzer/Documents/sdhmR-V2021.1/sdhmR-V2021.1EXP/Data/module02")
st_write(pied.range.fnetSF, dsn = ".", layer = "piedrange_fnetSF", driver = "ESRI Shapefile",
         delete_layer = T, delete_dsn = T)

setwd("C:/Users/epolzer/Documents/sdhmR-V2021.1/sdhmR-V2021.1EXP/Data/module02")
save("pied.range.fnetSF", "pied.range.fnetDF", file = "pied.range.fnet.RData")

setwd("C:/Users/epolzer/Documents/sdhmR-V2021.1/sdhmR-V2021.1EXP/Data/module02")
save(pied.indexFNET, file = "pied.indexFNET.RData")

setwd("C:/Users/epolzer/Documents/sdhmR-V2021.1/sdhmR-V2021.1EXP/Data/module02")
write.csv(pied.PPsA, file = "pied_PPsA.csv", row.names = F)
save(pied.PPsA, file = "pied.PPsA.RData")
```
---

## The End

---
