---
title: "Building Training Dataset"
author: "Eli L. Polzer"
date: "25 February 2021"
output: html_document
---
--> 
```{r}
path.root <- "C:/Users/epolzer/Documents/sdhmR-V2021.1"
path.dat <- paste(path.root, "/sdhmR-V2021.1EXP/Data/exercise/traindat", sep = "")
path.mod2 <- paste(path.root, "/sdhmR-V2021.1EXP/Data/module02", sep = "")
path.gis <- paste(path.root, "/sdhmR-V2021.1EXP/Data/gis_layers", sep = "")
path.figs <- paste(path.root, "/sdhmR-V2021.1EXP/Figures", sep = "")
path.prj <- paste(path.root, "/sdhmR-V2021.1EXP/R_Code", sep = "")

## Source course CRSs:
setwd(path.prj)
source("prjs4_sdhmR.R") # source course CRSs
ls(pattern = "prj.")

library(raster)
library(sf)

## Load data:
setwd(path.dat)
tru.range <- read.csv("spp106pr_RANG_2.csv")

setwd(path.mod2)
range.PPsA <- get(load("pied.PPsA.RData"))

setwd(path.mod2)
range.bufR <- get(load("pied.framesR.RData"))
range.bufSF <- get(load("pied.framesSF.RData"))
range.bufSP <- get(load("pied.framesSP.RData"))
range.bufpt <- get(load("pied.framesbufpt.RData"))

range.indexFNET <- get(load("range.indexFNET.RData"))
range.ppsAx <- get(load("range_pres.psu.RData"))
p2.sppx <- get(load("range_pres.psu.RData"))
p2.modFRx <- get(load("range_pres.psu.RData"))
psu.bufptx <- get(load("range_pres.psu.RData"))
```

---
## Question #1

* Import the predictor GIS variables into R

```{r}
# Joint import of predictor GIS variables into R:
setwd("C:/Users/epolzer/Documents/sdhmR-V2021.1/sdhmR-V2021.1EXP/Data/exercise/preds")
env.list <- list.files(pattern = ".img$") 
env.list 
layers <- {} 
for (i in 1:length(env.list)) {
  r1 <- crop(raster(env.list[i]), range.bufptR) 
  names(r1) <- strsplit(env.list[i], "_wgs.img") 
  layers <- c(layers, r1)
}
layers

# Individual import:
etpt5 <- raster("etpt_5.img")
etpt6 <- raster("etpt_6.img")
etptsprin <- raster("etpt_sprin.img") 
exp1nrm <- raster("exp1nrm.img") 
exp3nrm <- raster("exp3nrm.img") 
exp5nrm <- raster("exp5nrm.img")
mindyrav <- raster("mind_yr_av.img")
pradswdi <- raster("prad_sw_di.img") 
precwhal <- raster("prec_w_hal.img") 
precwinte <- raster("prec_winte.img") 
taveshal <- raster("tave_s_hal.img") 
tavesprin <- raster("tave_sprin.img") 
tmaxshal <- raster("tmax_s_hal.img") 
tmaxsumme <- raster("tmax_summe.img") 
rough <- raster("rough_1k.img") 
topos <- raster("topos.img") 
```

---

## Question #2

* Extract predictor variables from all 16 **`.img`** files using the presence:absence [X,Y]'s
```{r}
 # Individual extract:
setwd(path.mod2)
range.bufR <- get(load("pied.framesR.RData"))
range.bufSF <- get(load("pied.framesSF.RData"))
range.bufSP <- get(load("pied.framesSP.RData"))
range.bufpt <- get(load("pied.framesbufpt.RData"))

range.indexFNET <- get(load("range.indexFNET.RData"))
range.ppsAx <- get(load("range_pres.psu.RData"))
p2.sppx <- get(load("range_pres.psu.RData"))
p2.modFRx <- get(load("range_pres.psu.RData"))
psu.bufptx <- get(load("range_pres.psu.RData"))

et1 <- extract(etpt5, range.PPsA[, c("tr.wgs_x", "tr.wgs_y")])   
length(et1)  
length(which(is.na(et1)))

et2 <- extract(etpt6, range.PPsA[, c("tr.wgs_x", "tr.wgs_y")])   
length(et2)  
length(which(is.na(et1)))

et3 <- extract(etptsprin, range.PPsA[, c("tr.wgs_x", "tr.wgs_y")])   
length(et3)  
length(which(is.na(et1)))

ex1 <- extract(exp1nrm, range.PPsA[, c("tr.wgs_x", "tr.wgs_y")])   
length(ex1)  
length(which(is.na(et1)))

ex3 <- extract(exp3nrm, range.PPsA[, c("tr.wgs_x", "tr.wgs_y")])   
length(ex3)  
length(which(is.na(et1)))

ex5 <- extract(exp5nrm, range.PPsA[, c("tr.wgs_x", "tr.wgs_y")])   
length(ex5)  
length(which(is.na(et1)))

mi1 <- extract(mindyrav, range.PPsA[, c("tr.wgs_x", "tr.wgs_y")])   
length(mi1)  
length(which(is.na(et1)))

pd1 <- extract(pradswdi, range.PPsA[, c("tr.wgs_x", "tr.wgs_y")])   
length(pd1)  
length(which(is.na(et1)))

pe1 <- extract(precwhal, range.PPsA[, c("tr.wgs_x", "tr.wgs_y")]) 
length(pe1) 
length(which(is.na(et1)))

pe2 <- extract(precwinte, range.PPsA[, c("tr.wgs_x", "tr.wgs_y")])
length(pe2) 
length(which(is.na(et1)))

ta1 <- extract(taveshal, range.PPsA[, c("tr.wgs_x", "tr.wgs_y")]) 
length(ta1) 
length(which(is.na(et1)))

ta2 <- extract(tavesprin, range.PPsA[, c("tr.wgs_x", "tr.wgs_y")])  
length(ta2) 
length(which(is.na(et1)))

tm1 <- extract(tmaxshal, range.PPsA[, c("tr.wgs_x", "tr.wgs_y")])
length(tm1) 
length(which(is.na(et1)))

tm2 <- extract(tmaxsumme, range.PPsA[, c("tr.wgs_x", "tr.wgs_y")])
length(tm2) 
length(which(is.na(et1)))

ro1 <- extract(rough, range.PPsA[, c("tr.wgs_x", "tr.wgs_y")]) 
length(ro1) 
length(which(is.na(et1)))

to1 <- extract(topos, range.PPsA[, c("tr.wgs_x", "tr.wgs_y")]) 
length(to1) 
length(which(is.na(et1)))

# Joint extract:
range.var <- stack(layers) 
range.var 
d1 <- extract(range.var, range.PPsA[, c("tr.wgs_x", "tr.wgs_y")]) 
head(d1, 2) 
length(which(is.na(d1)))
```

---

## Question #3

* Create a new dataframe by binding the extracted predictor variable values to the dataframe from exercise #4.
```{r}
# Individual bind: 

range.train.ind <- cbind(range.PPsA, et1, et2, et3, ex1, ex3, ex5, mi1, pd1, pe1, pe2, ta1, ta2, tm1, tm2, ro1, to1)
head(range.train.ind, 2)
length(range.train.ind)

# Joint/Stacked bind:
d1 <- extract(range.var, range.PPsA[, c("tr.wgs_x", "tr.wgs_y")]) 
head(d1, 2) 
range.train <- cbind(range.PPsA, d1)
head(range.train, 2) 
length(range.train)
```

---

## Question #4

* Save your data as R objects:
  * Dataframe;
  * Point shapefile with geometry in R
  * Export as a point shapefile in ESRI format
* Save these R objects in a **`.RData`** file
```{r}
setwd(path.mod2)
write.csv(range.train, file = "range.train.csv", row.names = F)
write.csv(range.train.ind, file = "range.train.ind.csv", row.names = F)
save(range.train, file = "range.train.RData")
save(range.train.ind, file = "range.train.ind.RData")
```

---

## The End

---
