---
title: "Explore the Training Dataset"
author: "Eli L. Polzer"
date: "28 February 2021"
output: html_document
---

## The Data

* The dataframe completed in exercise #5.

---

## Question #1

* Break the predictor variables into logical groups of **topography**, **temperature**, and **precipitation**.  Do this by creating a new column called **`pred_type`** and add it the dataframe.

```{r}
path.root <- "C:/Users/epolzer/Documents/sdhmR-V2021.1"
path.dat <- paste(path.root, "/sdhmR-V2021.1EXP/Data/exercise/traindat", sep = "")
path.mod2 <- paste(path.root, "/sdhmR-V2021.1EXP/Data/module02", sep = "")
path.gis <- paste(path.root, "/sdhmR-V2021.1EXP/Data/gis_layers", sep = "")
path.figs <- paste(path.root, "/sdhmR-V2021.1EXP/Figures", sep = "")
path.prj <- paste(path.root, "/sdhmR-V2021.1EXP/R_Code", sep = "")

library(modeest)
library(tidyverse) 
library(data.table)
library(sf)

## Load data:
setwd(path.mod2)
range.tr <- get(load("range.train.RData "))
dim(range.tr)
head(range.tr [c(1:2, 7:24)], 2)

## Group data:
pred_groups <- data.frame(names(range.tr[,9:24]))
names(pred_groups) <- "PredVariable"
topo <- c("exp1nrm.img","exp3nrm.img","exp5nrm.img","prad_sw_di.img","rough_1k.img","topos.img")
temp <- c("tave_s_hal.img","tave_sprin.img","tmax_s_hal.img","tmax_summe.img")
moisture <- c("etpt_5.img","etpt_6.img","etpt_sprin.img","mind_yr_av.img","prec_w_hal.img","prec_winte.img")

pred_groups$PredType <- ifelse(pred_groups$PredVariable %in% topo,"Topography",
                               ifelse(pred_groups$PredVariable %in% temp,"Temperature", "Moisture"))
```

---

## Question #2

* Assess correlations:
  * Among all variables; and
  * Within each of the logical groupings created above
```{r}
## Assess correlations of all individual variables:
cut.point <- 0.7
c1 <- cor(range.tr [, c(9:24)], use = "pairwise.complete.obs", method = "spearman") 
c1 

c2 <- subset(c1 > cut.point | c1 < cut.point )
c2 

## Assess correlations of grouped variables:
c3 <- cor(dplyr::select(range.tr, one_of(topo)), use = "pairwise.complete.obs", method = "spearman")
c_topo <- subset(c3 > cut.point | c3 < -cut.point)
c_topo

c4 <- cor(dplyr::select(range.tr, one_of(temp)), use = "pairwise.complete.obs", method = "spearman")
c_temp <- subset(c4 > cut.point | c4 < -cut.point)
c_temp

c5 <- cor(dplyr::select(range.tr, one_of(moisture)), use = "pairwise.complete.obs", method = "spearman")
c_moisture <- subset(c5 > cut.point | c5 < -cut.point)
c_moisture
```

---

## Question #3

* Examine variable importance of each predictor variable as related to presence:absence using the process described in the Module.
```{r}
varimp.glm <- function(tr.spp, tr.var, pres, pf, pl) {
  tmp.mat <- matrix(ncol = 2, nrow = (pl - pf + 1))
  for (i in pf:pl) {
    tmp <- glm(tr.spp[, pres] ~ tr.var[, i] + I((tr.var[, i])^2), na.action = na.omit,
               family = binomial)
    tmp.mat[(i - pf + 1), 1] <- tmp$aic
    tmp.mat[(i - pf + 1), 2] <- (1 - (tmp$deviance/tmp$null.deviance))
  }
  return(tmp.mat)
} 

tr.vip <- range.tr[, c(2, 9:24)] 
pres <- 1 
v.start <- 2 
v.stop <- ncol(tr.vip) 
v.num <- v.stop - 1 
dev.fit <- varimp.glm(tr.vip, tr.vip, pres, v.start, v.stop) 
dev.fit <- data.frame(pred_groups, dev.fit) 
setnames(dev.fit, c("X1", "X2"), c("AIC", "AdjDeviance"))
dev.fit1 <- dev.fit[order(dev.fit$PredType),]
dev.fit$PredVariable <- factor(dev.fit$PredVariable, 
                           levels = dev.fit$Variable[order(dev.fit$PredType)])
dev.fit1
```

---

## Question #4

* Eliminate redundant variables with a goal of retaining 7-10 of the 16 available
* Justify your decision(s) to keep / remove variables using bullets
```{r}
remove = c("etpt_6.img", "etpt_sprin.img", "prec_winte.img", "exp3nrm.img", "exp5nrm.img", "topos.img", "tave_s_hal.img", "tave_sprin.img", "tmax_summe.img")
range.tr.final <- (select(range.tr, !one_of(remove)))

#exp3nrm: high correlation with both exp1nrm and exp5nrm plus low AdjDev
#exp5nrm: high correlation with both exp1nrm and exp3nrm plus low AdjDev
#topos: somewhat high correlation withexp3nrm and exp5nrm and very low AdjDev (choice between this var and exp1nrm, which were very close)
#etpt6: high correlation with etpt5 and mind_yr_av and low AdjDev (choice between this var and etpt 5, which were very similar)
#etpt_sprin: high correlation with etpt5, etpt6, and mind_yr_av plus low AdjDev
#prec_wint: high correlation with prec_w_hal plus very low AdjDev
#tave_s_hal: high correlation with all vars plus lowest AdjDev of Temp var
#tave_sprin: high correlation with all vars plus 2nd lowest AdjDev of Temp var
#tmax_summe: high correlation with all vars plus 3rd lowest AdjDev of Temp var
```

---

## Question #5

* Include the term **tr_MORT**, **tr_PERS**, **tr_RANG**, and **tr_SEED** somewhere in the final training data objects, depending on your data group.  Use of **tr_ByGroup** here means the data are now ready for model creation.
* Save your data as R objects:
  * Dataframe;
  * Point shapefile with geometry in R
  * Export as a point shapefile in ESRI format
* Save these R objects in a **`.RData`** file

These data will be used as we next begin the SDHM model constructions.
```{r}
setwd(path.mod2)
write.csv(range.tr.final, file = "tr_RANG.FINAL.csv", row.names = F) # save .csv - training data
save(range.tr.final, file = "tr_RANG.FINAL.RData") # save .RData - pres/abs plus stacked raster
st_write(range.tr.final, dsn = ".", layer = "tr_RANG.FINAL", driver = "ESRI Shapefile",
         delete_layer = T, delete_dsn = T) #point shapefile in ESRI format
```

---

## The End

---
