---
title: "sdhmR Homework 2_M2.3.4_Building a Fishnet"
author: "Eli L. Polzer"
date: "09 February 2021"
output: html_document
---

<!-- 

--> 
```{r global_options, include=FALSE}
knitr::opts_knit$set(root.dir = "C:/Users/epolzer/Documents/sdhmR-V2021.1/sdhmR-V2021.1EXP")
knitr::opts_chunk$set(warning=FALSE,error=TRUE,message=FALSE)
```

---

## This exercise links to Module 2.3.4

Submit the completed exercise to us at:

* t.edwards@usu.edu 
* eric_tsakiris@fws.gov 

---

## Context

The Uinta Basin hookless cactus (*Sclerocactus wetlandicus*) (SCWE) is listed as a threatened species.  Recovery goals require:
* An accurate range map;
* Comprehensive surveys on locations; and
* Other needs as well
[FWS ECOS](https://ecos.fws.gov/ecp/species/9037) has an expert opinion-based "blob" of an hypothetical SCWE modelling (ie, the "blob") domain.

## The Data

* Acquire from the ECOS site the SCWE "blob" shapefile (call it: scwe_ecospoly) 
* Import the **`scwe_pres.csv`** file of known SCWE presence points
* Data in: **`~\sdhmR-V2020.2\data\gis_layers`** and **`...\module02`**, respectively

---

## Question #1

* Import the ECOS--acquired SCWE "blob" as a spatial object (call it: **`scwe_ecospoly`**)
* Find and acquire the **`CRS epsg:42303`** and it's **`proj4string`** – this is the CRS for ALL subsequent analyses
* Set CRS of (i) the ECOS "blob" and (ii) the imported points to epsg:42303
* Plot the presence points relative to the ECOS-based polygon "blob"

```{r}
## Root directory:
path.root <- "~/sdhmR-V2021.1EXP"
path.mod2 <- paste(path.root, "C:/Users/epolzer/Documents/sdhmR-V2021.1/sdhmR-V2021.1EXP/Data/module02", sep = "")
path.gis <- paste(path.root, "C:/Users/epolzer/Documents/sdhmR-V2021.1/sdhmR-V2021.1EXP/Data/gis_layers", sep = "")
path.figs <- paste(path.root, "C:/Users/epolzer/Documents/sdhmR-V2021.1/sdhmR-V2021.1EXP/Figures", sep = "")

## Source course CRS:
setwd("~/sdhmR-V2021.1EXP")
source("C:/Users/epolzer/Documents/sdhmR-V2021.1/sdhmR-V2021.1EXP/R Code/prjs4_sdhmR.R") # source course CRSs
ls(pattern = "prj.")

library(raster)
library(sf)
################################################################################

## Import data points:
setwd("C:/Users/epolzer/Documents/sdhmR-V2021.1/sdhmR-V2021.1EXP/Data/module02")
scwe.pres <-  read.csv("scwe_pres.csv")

## Import shapefile: 
setwd("C:/Users/epolzer/Documents/sdhmR-V2021.1/sdhmR-V2021.1EXP/Data/gis_layers")
ecospoly <- st_read(dsn = ".", layer = "scwe_ecospoly")

## Convert to sf & assign input CRS: 
scwe.pres_01 <- st_as_sf(scwe.pres, coords = c("aeaN83_x", "aeaN83_y"), 
                          crs = prj.aeaN83, remove = F) # create spatial DF
ecospoly_01 <- st_as_sf(ecospoly, coords = c("aeaN83_x", "aeaN83_y"), 
                        crs = prj.aeaN83, remove = F) # create spatial DF)

## Build no-points bounding box (ie LAND TENURE-BASED):
## Import ECOS SCWE polygon & plot: 
setwd("C:/Users/epolzer/Documents/sdhmR-V2021.1/sdhmR-V2021.1EXP/Data/gis_layers")
scwe.noptSF <- st_read(dsn = ".", layer = "scwe_ecospoly")
scwe.noptSF # examine
plot(st_geometry(scwe.noptSF), axes = T)
plot(st_geometry(ecospoly_01), axes = T)

##Plot points in blob:
points(scwe.pres$aeaN83_x , scwe.pres$aeaN83_y , pch = 20, col = "darkgreen")
```
---

## Question #2

* Generate a buffered point-based (ie OPTION #4 from Mod 2.3.4) modelling frame
* Set buffer of 10,000 m
* Set CRS of modelling frame to epsg:42303

```{r}
## Obtain CRS for modelling frame & res from existing GIS data layer in aeaN83:
setwd("C:/Users/epolzer/Documents/sdhmR-V2021.1/sdhmR-V2021.1EXP/Data/gis_layers")
t1 <- raster("elev_1k_aea.img")
t1 # examine

## Buffer all pts by 10km; merge overlapping polys:
scwe.bufptSF <- st_buffer(scwe.pres_01, dist = 10000) # retains all polys
scwe.bufptSF <- st_union(scwe.bufptSF)  # merges all intersecting polys
scwe.bufptSF # examine

## Extract buffered points from template & rasterize:
scwe.bufptSP <- as_Spatial(scwe.bufptSF) 
scwe.bufptR <- crop(rasterize(scwe.bufptSP, t1), scwe.bufptSP) 
scwe.bufptR 
```
---

## Question #3

* Select larger of the two frames as FISHNET extent 
  * (HINT: examine extents of both -- one will be larger)

```{r}
plot(st_geometry(ecospoly_01), axes = T)
plot(scwe.bufptSF, add = T, border = "red") # add border to buffer pts
points(scwe.pres$aeaN83_x, scwe.pres$aeaN83_y , pch = 20, col = "darkgreen") # add points to plot
```

The frame created with the 10km-buffered points (seen in red) is slightly larger than the ECOS polygon blob frame, so suggest using the buffered point frame as the FISHNET extent. 
---

## Question #4

* Build a 1,000 m (1 km) square FISHNET
* Save both spatial and dataframe FISHNETs as R objects 
  * (NOTE: we will use these later so be sure and save !!!)
* Export as both dataframe (ie, no geometry) and ESRI shapefile

```{r}
## Obtain CRS for modelling frame & res from existing GIS data layer in aeaN83:
setwd("C:/Users/epolzer/Documents/sdhmR-V2021.1/sdhmR-V2021.1EXP/Data/gis_layers")
t1 <- raster("elev_1k_aea.img")
t1 # examine
  
## Buffer all pts by 1km; merge overlapping polys:
scwe.bufptSF.1 <- st_buffer(scwe.pres_01, dist = 1000) # retains all polys
scwe.bufptSF.1k <- st_union(scwe.bufptSF.1)  # merges all intersecting polys
scwe.bufptSF.1k # examine
#plot(st_geometry(scwe.bufptSF.1k), axes = T)# NOT RUN
  
## Extract buffered points from template & rasterize:
scwe.bufptSP.1k <- as_Spatial(scwe.bufptSF.1k) 
scwe.bufptR.1k <- crop(rasterize(scwe.bufptSP.1k, t1), scwe.bufptSP.1k) 
scwe.bufptR.1k 

## Load rasterized spatial frames:
setwd("C:/Users/epolzer/Documents/sdhmR-V2021.1/sdhmR-V2021.1EXP/Data/module02")
load("scwe.framesR.RData") # load raster frames if not in workspace

## Examine frames:
#scwe.bufptR # rectangle based on 1km buffer; spp point-based polygons

## Extract coords, cell id of coords, & build dataframe:
f1 <- sp::coordinates(scwe.bufptR.1k) # get spatial coords from pied.bufR
f2 <- cellFromXY(scwe.bufptR.1k, f1) # extract cell number from buffered extent
f3 <- as.data.frame(cbind(f1, f2)) # build datframe of x,y & cell number
head(f3, 2) # examine
tail(f3, 1) # f2 value should eq. ncell in pied.bufR; below
ncell(scwe.bufptR.1k) # should eq. f2 value above

## Clean up:
names(f3)[1:3] <- c("cell.aeaN83_x", "cell.aeaN83_y", "FNETID") # assign names
f3 <- f3[c("FNETID", "cell.aeaN83_x", "cell.aeaN83_y")] # reorder

## Examine fishnet data structure:
dim(f3) # dimension; rows is maximum number of cells in fishnet 
names(f3) # names in fishnet
head(f3, 2) # note first FNETID number
tail(f3, 2) # note last FNETID number matches dimension size

## Convert dataframe to spatial object:
scwe.fnetSF.1k <- st_as_sf(f3, coords = c("cell.aeaN83_x", "cell.aeaN83_y"), 
                        crs = prj.aeaN83, remove = F) # remove=F retains input x,y)
head(scwe.fnetSF.1k, 2) # examine

## Build stand-alone dataframe:
class(scwe.fnetSF.1k) # note dataframe class; will extract data.frame
scwe.fnetDF.1k <- st_drop_geometry(scwe.fnetSF.1k) # build dataframe
head(scwe.fnetDF.1k, 2) # examine
class(scwe.fnetDF.1k) # check if class=data.frame

## Export as ESRI shapefile & dataframe:
setwd("C:/Users/epolzer/Documents/sdhmR-V2021.1/sdhmR-V2021.1EXP/Data/module02") # output path
st_write(scwe.fnetSF.1k, dsn = ".", layer = "scwe_fnetSF", driver = "ESRI Shapefile",
         delete_layer = T, delete_dsn = T) # output shapefile

## Save objects:
setwd("C:/Users/epolzer/Documents/sdhmR-V2021.1/sdhmR-V2021.1EXP/Data/module02") # output path
save("scwe.fnetSF.1k", "scwe.fnetDF.1k", file = "scwe.fnet.1k.RData")
```
---

## Question #5

Once you have completed the analyses above, answer the following:

* How do the known SCWE presences relate to the modelling frame "blob"?

* How different is the SCWE "blob" from a buffered point-based modelling domain?

* What is the best "blob" for creating a data indexing FISHNET?  Why?

---

## The End

---
