---
title: "HW 7: Logistic Regression"
author: "Group A: Eli Polzer, Matt Futia, Mia McReynolds"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: html_document
---

```{r global_options, include=FALSE}
# knitr::opts_knit$set(root.dir = "~/University of Vermont/Classes/sdhmR/sdhmR-V2021.1")
knitr::opts_chunk$set(warning=FALSE,error=TRUE,message=FALSE)
```

```{r}
# some pathnames; yours will be specific to your CPU !!
  path.root <- "~/University of Vermont/Classes/sdhmR/sdhmR-V2021.1" 
  
# below is the recommended class root dir
  #path.root <- "~/sdhmR-V2021.1"  # typical class root dir
  path.mod2 <- paste(path.root, "/data/module02", sep = "")
   path.mod5 <- paste(path.root, "/data/module05.01-LR", sep = "")
  path.figs <- paste(path.root, "/powerpoints/figures", sep = "") 
  path.gis <- paste(path.root, "/data/gis_layers", sep = "")
  
# some libraries
  library(tidyverse)
   library(PresenceAbsence)  # fxns: optimal.thresholds, presence.absence.accuracy, 
                            #       auc.roc.plot
  library(DAAG)             # fxns: CVbinary
  library(data.table)
  library(raster)
  library(sf)
```


## Question #1

* Import and explore data
  * **NOTE**:  intent is not to reduce number of variables; that was completed in exercise #5.  Rather, calculate simple descriptive statistics (mean, sd, n) and boxplots


```{r}
# load range-wide training data
load("~/University of Vermont/Classes/sdhmR/sdhmR-V2021.1/data/module02/tr_RANG.Rdata")

str(tr_RANG)
```

```{r}
#pivot for ggplotting
tr_RANG_long <- tr_RANG %>% 
  pivot_longer(etpt_5:topos, names_to = "Variable", values_to = "Value") 

head(tr_RANG_long, 3)
```


```{r}
#summarise data 
tr_RANG_long %>% group_by(Variable) %>% 
  summarise(n = n(),
            mean = mean(Value), 
            sd = sd(Value), 
            min = min(Value), 
            max = max(Value))
```
```{r}
#boxplot
ggplot(tr_RANG_long, aes(y=Value)) +
  geom_boxplot() +
  facet_wrap(~Variable, scales="free") +
  theme_classic()
```


---

## Question #2

* Construct full- and reduced-variable logistic GLM’s
* Determine an estimate of model fit, based on deviance.


```{r}
######## START MODEL #1 => FULL MODEL, ALL VARS INCLUDED
  mod1.LR.RANG <- glm(as.factor(RANG106) ~ etpt_5 + mind_yr_av + prad_sw_di + prec_w_hal + rough_1k + tmax_s_hal + topos, family = binomial, data = tr_RANG)

# model 1 summary
  summary(mod1.LR.RANG) # full model summary stats 
```

```{r}
# model 1 fit
  mod1.fit <- 100 * (1 - mod1.LR.RANG$deviance/mod1.LR.RANG$null.deviance) 
      # extract values from above
  mod1.fit  # examine fit
```

```{r}
# model 1 prediction
mod1.pred <- predict(mod1.LR.RANG, type = "response") # model prediction
head(mod1.pred) # examine prediction - should be same length as input dataframe
```

In the full model, all variables were at least moderately significant. However, model fit was poor-- only 12%.


```{r}
######## START MODEL #2 => REDUCED VARIABLE MODEL
# build parsimonious model w/var reduction techniques: mod2.LR
#   variable reduction: backwards
  mod2.LR.RANG <- step(mod1.LR.RANG, trace = F) # backwards stepwise variable reduction
  summary(mod2.LR.RANG) # reduced model 
```

```{r}
# model 1 v. model 2 fit: both pretty bad
  100 * (1 - mod1.LR.RANG$deviance/mod1.LR.RANG$null.deviance)  # fit model 1
  100 * (1 - mod2.LR.RANG$deviance/mod2.LR.RANG$null.deviance)  # fit model 2
```
Both of these models have the same formula, so their fit is the same. Not very good overall. 

```{r}
# model 1 vs. 2 predictions
  mod2.pred <- predict(mod2.LR.RANG, type = "response") # model prediction
  head(mod2.pred)
  head(mod1.pred)
```


---

## Question #3

* Calculate accuracy metrics (full set as in Module 3.2 and 3.3, Analytical Intermission) using:
  * Resubstitution approaches, and
  * A 10-fold cross--validation approach
  
### Resubstitution approaches

```{r}
######## START RESUBSTITUTION ACCURACY CALCULATIONS, MODEL=LOGISTIC GLM
# requires pkg PresenceAbsence
#   build testing dataframe using mod2 predictions
  modl <- "mod.LR.RANG" # add var to keep track of model
  dat2 <- cbind(modl, tr_RANG[2], mod2.pred) # build dataframe w/mod2 predictions
  head(dat2, 2) # examine prediction dataframe
```


```{r}
# determine best threshold using PresenceAbsence package Sec7.1
  mod.cut <- optimal.thresholds(dat2, opt.methods = c("Default")) # default threshold=0.5
  mod.cut # examine threshold=DEFAULT of 0.5
```

```{r}
# generate confusion matrix
  mod2.cfmat <- table(dat2[[2]], factor(as.numeric(dat2$mod2.pred >= mod.cut$mod2.pred)))
  mod2.cfmat # examine
```

```{r}

# calculate model accuracies with standard deviation=F
  mod2.acc <- presence.absence.accuracy(dat2, threshold = mod.cut$mod2.pred, st.dev = F)
  tss <- mod2.acc$sensitivity + mod2.acc$specificity - 1 # code TSS metric
  mod2.acc <- cbind(mod2.acc[1:7], tss) # bind all metrics
  mod2.acc[c(1, 4:5, 7:8)] # examine accuracies

```

```{r}
# plotting AUC
  auc.roc.plot(dat2, color = T) # basic AUC plot; pkg PresenceAbsence 
```

### 10-fold cross validation approach

```{r}
# perform 10-fold cross-validation
  set.seed(1234) # set and save seed if desire replicability of samples
  #library(DAAG) # load pkg for crossval
  glm.cv10 <- CVbinary(mod1.LR.RANG, nfolds = 10, print.details = F) # crossval w/10 folds
  head(glm.cv10$cvhat) # examine
```
### Compare both

```{r}
# examine resubstitution and cross-validation estimates
  glm.pred <- mod2.pred # examine resub prediction 
head(glm.pred)
head(glm.cv10$cvhat) # examine xval prediction
  glm.cvpred <- glm.cv10$cvhat # assign new name to crossval estimates
  spp <- "spp106"
  dat2 <- cbind(spp, tr_RANG[2], glm.pred, glm.cvpred) # build dataframe: true data plus the 2 validations
  head(dat2, 2) # examine; NOTE will differ each run unless seed is saved 
  
```

```{r}
# estimate both resub and crossval accuracies
#  create 3-column dataframes for PresenceAbsence
  dat3 <- dat2[, c(1:3)] # includes column glm.pred
  dat4 <- dat2[, c(1:2, 4)] # includes column glm.cvpred
```


```{r}
# calculate model accuracies; resubstitution
  mod.cut <- optimal.thresholds(dat3, opt.methods = c("ObsPrev")) # threshold=PREVALENCE
  mod1.acc <- presence.absence.accuracy(dat3,threshold = mod.cut$glm.pred)
  tss <- mod1.acc$sensitivity + mod1.acc$specificity - 1 # code TSS metric
  dat3.acc <- cbind(mod1.acc[1:7], tss) # bind all metrics
  head(dat3.acc) # examine
```

```{r}

# calculate model accuracies; cross-validation
  mod.cut <- optimal.thresholds(dat4, opt.methods = c("ObsPrev")) # threshold=PREVALENCE
  mod1.acc <- presence.absence.accuracy(dat4, threshold = mod.cut$glm.cvpred)
  tss <- mod1.acc$sensitivity + mod1.acc$specificity - 1 # code TSS metric
  dat4.acc <- cbind(mod1.acc[1:7], tss) # bind all metrics
  head(dat4.acc) # examine

```


```{r}
# bind all accuracies into single dataframe
  all.acc <- rbind(dat3.acc, dat4.acc) # bind all metrics
  all.acc # examine - usually just report the cvpred (cross-validated).

```

Very similar results in resubstitution and cross-validation approaches, so this model may be able to handle new data decently.

---

## Question #4

* Build 2 prediction maps:
  * A raw probability estimate for each cell in the modelling domain; and
  * A classified map based on the selected threshold from Question #3
  
```{r raw probability estimate}
#load raster stack - pied.topoDOM
load("~/University of Vermont/Classes/sdhmR/sdhmR-V2021.1/data/module02/rangeDOM.RData") 

# predict for LR & GAM
#  S=raster stack; M=RF model; N.img=save an .img file
#  accept the rest as R catechism specific to LR & GAM
rang.probLR <- predict(rangeDOM,mod1.LR.RANG,filename="rang.probLR.img",type="response",fun=predict,index=2, overwrite=TRUE)
rang.probLR
```
  
```{r classified map data}
# classification raster; bit more complicated 
#   first determine classification threshold -- did this above.
  mod.cut 

rang.clasLR <- reclassify(rang.probLR, filename = "range.clas.LR.img", 
   (c(0, mod.cut[[2]], 0, mod.cut[[2]], 1, 1)), overwrite=TRUE)
```
```{r plot maps}
######## START PROBABILITY & CLASSIFICATION PLOTS	 
# boundaries for pretty maps (see Mod 2.5)
  setwd(path.gis)
  states <- st_read(dsn = ".", layer = "na_states_wgs") # import shapefile
```

```{r}
# giggle maps
  par(mfrow = c(1, 2))
  plot(rang.probLR, legend = T, axes = T, main = "Probability Map") # plot probability map
  plot(st_geometry(states), add = T, lwd = 1.5) # add state boundaries
  plot(rang.clasLR, legend = T, axes = T, main = "Classification Map") # plot classification map
  plot(st_geometry(states), add = T, lwd = 1.5) # add state boundaries
  par(mfrow = c(1, 1))
  
  LRmaps_EX07 <- recordPlot()
```


---

## Question #5

* Save your data as R objects:
  * Accuracy metrics as a dataframe;
  * Classification threshold as a stand--alone scalar
  * Both prediction maps as **`.img`** format
* Save these R objects in a **`.RData`** file

These data will be used again in Module 10, Ensemble Models.

```{r}
setwd(path.mod5)
# Accuracy metrics as a dataframe
all.acc_EX07 <- all.acc

# Classification threshold as a stand--alone scalar
mod.cut_EX07 <- mod.cut[[2]]

#plot - also saved this manually
LRmaps_EX07

save(all.acc_EX07, mod.cut_EX07, LRmaps_EX07, file = "range.LR_EX07.RData")
```

---

## The End

---
