---
title: "Exercise Title Goes Here"
author: "Your Name Goes Here"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: html_document
---

<!-- These are global options  
       set root directory here 
       NOTE mine is active; yours is commented out
       change the path to reflect your CPU,
       then comment out or delete mine
--> 
```{r global_options, include=FALSE}
#knitr::opts_knit$set(root.dir = "~/words/classes/sdmR_ALLversions/sdhmR-V2020.2")
knitr::opts_knit$set(root.dir = getwd())
knitr::opts_chunk$set(warning=FALSE,error=TRUE,message=FALSE)

#filepaths
  path.root = getwd() 
  path.mod2 = paste(path.root, "/data/module02", sep = "")
  path.figs = paste(path.root, "/powerpoints/figures", sep = "") 
  path.gis = paste(path.root, "/data/gis_layers", sep = "")
```

---

## This exercise links to Module 2.3.4

Submit the completed exercise to us at:

* t.edwards@usu.edu 
* eric_tsakiris@fws.gov 

---

## Context

The Uinta Basin hookless cactus (*Sclerocactus wetlandicus*) (SCWE) is listed as a threatened species.  Recovery goals require:
* An accurate range map;
* Comprehensive surveys on locations; and
* Other needs as well
[FWS ECOS](https://ecos.fws.gov/ecp/species/9037) has an expert opinion-based "blob" of an hypothetical SCWE modelling (ie, the "blob") domain.

## The Data

* Acquire from the ECOS site the SCWE "blob" shapefile (call it: scwe_ecospoly) 
* Import the **`scwe_pres.csv`** file of known SCWE presence points
* Data in: **`~\sdhmR-V2020.2\data\gis_layers`** and **`...\module02`**, respectively
```{r}
```

---

## Question #1

* Import the ECOS--acquired SCWE "blob" as a spatial object (call it: **`scwe_ecospoly`**)
* Find and acquire the **`CRS epsg:42303`** and it's **`proj4string`** – this is the CRS for ALL subsequent analyses
* Set CRS of (i) the ECOS "blob" and (ii) the imported points to epsg:42303
* Plot the presence points relative to the ECOS-based polygon "blob"

```{r}
library(raster)
library(sf)
library(foreign)
library(magrittr)
library(ggplot2)

scwe_ecospoly = st_read("data/gis_layers/scwe_ecospoly.shp") # SCWE shp (polygon)
scwe_p = read.csv("data/module02/scwe_pres.csv")             # SCWE presence points 
scwe_p$date.r %<>% as.Date(., format = "%Y-%m-%d")

prj_epsg42303 = "+proj=aea +lat_1=29.5 +lat_2=45.5 +lat_0=23 +lon_0=-96 +x_0=0 +y_0=0 +ellps=GRS80 +datum=NAD83 +units=m +no_defs no_defs"

prj_aeaN83 = "+proj=aea +lat_1=29.5 +lat_2=45.5 +lat_0=23 +lon_0=-96 +x_0=0 +y_0=0 +ellps=GRS80 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs" # epsg:42303 
```

```{r}
# transform ECOS blob
scwe_ecospoly.1 <- st_transform(scwe_ecospoly, crs = prj_epsg42303)
st_crs(scwe_ecospoly.1)$proj4string

# transform imported points
scwe_points = st_as_sf(scwe_p, coords = c("aeaN83_x","aeaN83_y"), crs = prj_aeaN83, remove = F) # convert to point shapefile
scwe_points = st_transform(scwe_points, crs = prj_epsg42303)
st_crs(scwe_points)$proj4string

#extract points with new CRS and add to original dataframe
coords = data.frame(st_coordinates(scwe_points))
names(coords) = c("NAD83_x", "NAD83_y")
scwe_points = st_sf(data.frame(scwe_points, coords))

```

```{r}
ggplot(scwe_ecospoly.1)+
  geom_sf(color = "gray95")+
  geom_point(data = scwe_points, aes(aeaN83_x.1, aeaN83_y.1), shape = 1, color = "green", inherit.aes = F)+
  theme_classic()

```

---

## Question #2

* Generate a buffered point-based (ie OPTION #4 from Mod 2.3.4) modelling frame
* Set buffer of 10,000 m (10 km = .1 degrees)
* Set CRS of modelling frame to epsg:42303

```{r}
t1 <- raster("data/gis_layers/elev_1k_aea.img") # get base raster for conversion of polygon (sf)
# ??? How did you pick this one; do you look at all until find one with epsg:42303???

scwe_buf.points = st_buffer(scwe_points, dist = 10000) # create buffer of set distance
# ??? How do you know units of dist???

scwe_buf.points <- st_union(scwe_buf.points)  # merges all intersecting polys
scwe_buf.points # examine

scwe_buf_sp = as_Spatial(scwe_buf.points) #convert SF polygon to sp
scwe.buf_R = crop(rasterize(scwe_buf_sp, t1), scwe_buf_sp) # convert to raster and crop raster layer to extent of original geospatial frame

crs(scwe.buf_R)
```

---

## Question #3

* Select larger of the two frames as FISHNET extent 
  * (HINT: examine extents of both -- one will be larger)
  
```{r}
scwe_buf.points #bbox: -1205096, -1109030, 1933252, 2005708 (xmin, xmax, ymin, ymax)
scwe_ecospoly.1 #bbox: -1206694, -1112295, 1931294, 2003811

-1205096 - (-1109030) #96066
-1206694 - (-1112295) #94399

2005708 - 1933252 #72456
2003811 - 1931294 #72517

ggplot()+
  geom_sf(data = scwe_buf.points, color = "black", fill = NA) +
  geom_sf(data = scwe_ecospoly.1, color = "red", fill = NA)+
  geom_point(data = scwe_points, aes(aeaN83_x.1, aeaN83_y.1), shape = 1, color = "green")+
  theme_classic()

```

The frame made from the buffered points is larger (larger extent)

---

## Question #4

* Build a 1,000 m (1 km) square FISHNET
* Save both spatial and dataframe FISHNETs as R objects 
  * (NOTE: we will use these later so be sure and save !!!)
* Export as both dataframe (ie, no geometry) and ESRI shapefile

```{r}
f1 = sp::coordinates(scwe.buf_R) # get spatial coords from pied.bufR
f2 = cellFromXY(scwe.buf_R, f1) # extract cell number from buffered extent (create fishnet id)
f3 = as.data.frame(cbind(f1, f2))

names(f3) = c("cell.wgs_x", "cell.wgs_y", "FNETID")
  
scwe.fnetSF = st_as_sf(f3, coords = c("cell.wgs_x", "cell.wgs_y"), crs = prj_epsg42303, remove = F) # build spatial frame
scwe.fnetDF = st_drop_geometry(scwe.fnetSF) # build dataframe
```

```{r}
# save objects
save("scwe.fnetSF", "scwe.fnetDF", file = "swce.fnet.RData") # save spatial and df fishnets

st_write(scwe.fnetSF, dsn = ".", layer = "scwe_fnetSF", driver = "ESRI Shapefile", delete_layer = T, delete_dsn = T) # output shapefile
```

---

## Question #5

Once you have completed the analyses above, answer the following:

* How do the known SCWE presences relate to the modelling frame "blob"?
* How different is the SCWE "blob" from a buffered point-based modelling domain?
* What is the best "blob" for creating a data indexing FISHNET?  Why?

*Known presences all fall within the blob*

```{r}
((st_area(scwe_buf.points) - st_area(scwe_ecospoly.1))/st_area(scwe_ecospoly.1))*100
```

*The differences in area between the two frames is about 10%*

*The buffered point-based frame would be better because it has a larger extent and therefore may incorporate more environmental variables*
---

## The End

---
