---
title: "Exercise Title Goes Here"
author: "Your Name Goes Here"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: html_document
---

<!-- These are global options  
       set root directory here 
       NOTE mine is active; yours is commented out
       change the path to reflect your CPU,
       then comment out or delete mine
--> 
```{r global_options, include=FALSE}
knitr::opts_knit$set(root.dir = "~/words/classes/sdmR_ALLversions/sdhmR-V2020.2")
#knitr::opts_knit$set(root.dir = "~/sdhmR-V2020.2")
knitr::opts_chunk$set(warning=FALSE,error=TRUE,message=FALSE)
```

---

## This exercise links to Module 2.4

Submit the completed exercise to us at:

* t.edwards@usu.edu 
* eric_tsakiris@fws.gov  

---

## Context

The data are of the *Pinus edulis* Common piñon, the species you dealt with in Exercise #3.  In this exercise we assume the acquisition of species location data reveals that data are presence-only (measured 1's only; no 0=absence)

The goal is to build a dataframe consisting of pseudo--absences selected from *edulis*' geographic and environmental background.

### The Twist to this Exercise

There's a twist to what you will do in exercises starting now, with all subsequent exercises being group--based.  You can self--organize into groups, or I can assign you to a group.  Groups are on the order of 2--4 individuals. 

* **IMPORTANT**: If you wish to self--organize do that quickly and let Eric and I know ASAP.  You will need to provide the name and email address of those in your group.  Pick a leader to act as the principal POC.

Each group will be assigned one type of data, as descried below.  Once assigned a type of data, you will use those data for the rest of the class.

The data for *edulis*  have four discrete labels:

* **Seedlings**:  these data indicate spatial locations where seedlings of the tree have been found
* **Mortality**: these data indicate spatial locations where mortality has been observed
* **Persistence**: these data have no observed mortality or seedling at given spatial locations
* **Range**: the total dataset, including all spatial locations of seedlings, mortality, and persistence

---

## The Data

* Common piñon *Pinus edulis* sppcode=106
* Data for each group in:
  * **`~\sdhmR-V2020.2\data\exercise\traindat`** 
* Data files:
  * **`spp106pr_MORT.csv`**, **`spp106pr_PERS.csv`**, **`spp106pr_RANG.csv`**, **`spp106pr_SEED.csv`**

---

## Question #1

* Import the data type you have been assigned to work with
* Build a buffered-presence points bounding box
  * Buffer each point by ~50km 
  * **CAUTION**:  make sure the pluses and minuses when buffering are in the correct direction given north, south, east, and west !! (Serious screw-ups happen here …)
* Ensure you assign correct CRS, values, and resolution to bounding box

```{r}
# code for the question goes inside chunks, if part of the exercise solution.
#  If not part of the exercise questions delete this.
#  NOTE: do NOT answer all questions from within the same chunk
```

**YET ANOTHER NOTE RELATED TO CODE:**  
If King's English interpretation is required, it will go here, after each code chunk.

---

## Question #2

* Build a fishnet for the bounding box

---

## Question #3

* Use the fishnet to create a pseudo--absence data frame of [X,Y]’s
* For simplicity’s sake, set $n$ for pseudo--absence at twice (2$\times$) the number of presences in your respective RANG, PERS, MORT, and SEED data set
* Bind these data to the presence dataframe you imported in Question #1

---

## Question #4

* Save your data as R objects, including both the true presence and pseudo-absences:
  * Dataframe;
  * Point shapefile with geometry in R
  * Export as a point shapefile in ESRI format
* Save these R objects in a **`.RData`** file
* Save the bounding box as a raster--based **`.img`** file

---

## The End

---
