---
title: "Exercise Title Goes Here"
author: Eli Polzer, Mia McReynolds, Matt Futia
date: "`r format(Sys.time(), '%d %B %Y')`"
output: html_document
---

<!-- These are global options  
       set root directory here 
       NOTE mine is active; yours is commented out
       change the path to reflect your CPU,
       then comment out or delete mine
--> 
```{r global_options, include=FALSE}
knitr::opts_knit$set(root.dir = "~/words/classes/sdmR_ALLversions/sdhmR-V2020.2")
#knitr::opts_knit$set(root.dir = "~/sdhmR-V2020.2")
knitr::opts_chunk$set(warning=FALSE,error=TRUE,message=FALSE)
```

---

## This exercise links to Module 2.3.3

Submit the completed exercise to us at:

* t.edwards@usu.edu 
* eric_tsakiris@fws.gov 

---

## Context

The exercise explores the consequences of frame selection on numerical summaries of predictive variables.  The species being explored is [*Pinus edulis* Common piñon](https://www.conifers.org/pi/Pinus_edulis.php).  Feel free to refer to the illustrations in Module 2.3.2 for graphical context.

## The Data

* Load **`spp106.soil.Rdata`** and **`spp106.clim.Rdata`** (or import **`spp106.soil.csv`**, **`spp106.clim.csv`**)
* Data in: **`~\sdhmR-V2020.2\data\module02`**

---

## Question #1

* What are the expected consequences in terms of a SDHM classification model if differences between environmental variables measured at presence:absence sites are large vs. small?

```{r}
# code for the question goes inside chunks, if part of the exercise solution.
#  If not part of the exercise questions delete this.
#  NOTE: do NOT answer all questions from within the same chunk
```

**YET ANOTHER NOTE RELATED TO CODE:**  
If King's English interpretation is required, it will go here, after each code chunk.

---

## Question #2

* How might these differing frames, and their respective variable differences based on the presence or absence of the species, affect: 
  * (i) ecological interpretations, and 
  * (ii) their classification ability, of variables?

---

## Question #3

* How do you decide which frame -- given many competing -- is appropriate for SDHM construction?

---

## Question #4

* Examine variable names and select 3 for all subsequent questions.  Identify them below in your answer to this question.

---

## Question #5

* Run some simple descriptive statistics (ie, $\bar{x}$, $SD$) on the three (3) selected variables from above and compare across frames (where 1=inside frame; 0=outside): 
* There are four (4) frames to evaluate:
  * Western U.S. states; this frame is defined as the entire dataframe N
  * **`$dry.domfr`** = all *edulis* within the dry domain
  * **`$SPPRES106.fiafr`** = buffered *edulis* distribution based on FIA data points
  * **`$SPPRES106.cmbfr`** = buffered *edulis* distribution based on FIA data points plus the UNION with [Little’s mapped tree distributions](https://www.fs.fed.us/database/feis/pdfs/Little/aa_SupportingFiles/LittleMaps.html).
  
---
  
## Question #6

* Build box plots with the 4 frames by variable for each of your selected variable, ie this will be 3 variables $\times$ 4 different frames.

---

## Question #7: 

* Identify which of the variables that subjectively (No statistical tests here, please!Just your Mark I eyeballs.) look as though they would discriminate between presence and absence

---

## The End

---