---
title: "Exercise 8 Group A"
author: "Mia McReynolds, Eli Polzer, Matt Futia"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: html_document
---

```{r global_options, include=FALSE}
knitr::opts_knit$set(root.dir = "C:/Users/mttft/Desktop/UVM/UVM Courses/SDHM_R/sdhmR-V2021.1EXP")
knitr::opts_chunk$set(warning=FALSE,error=TRUE,message=FALSE)
```

---

## This exercise links to Module 5.2

Submit the completed exercise to us at:

* t.edwards@usu.edu 
* eric_tsakiris@fws.gov  

---

## Context

The data are of the *Pinus edulis* Common piñon, the species you dealt with in Exercise #3 -- #5.  This exercise relies on the dataframe you generated in exercise #6.  Hopefully you saved those combined presence:absence dataframe, with the reduced variables you begin SDHM model construction with, or regrettably you will need to return to exercise #6 and re--build that dataframe.

The goal is to:
* Variables in final model, significance, and direction of relationship
  * List which variables are "smoothed" vs, those that are not
* Plots of the "smoothed" variables
* Calculate an estimate of model fit
* Build a table of accuracy metrics, and AUC plot
* Provide some bulleted interpretation points
* Plot map products of the:
  * (i) probability; and 
  * (ii) classified distribution model
  * (Save these two maps as **`.img`** files for use later in the course)


Remember, you will have been assigned data for *edulis* having one the following four discrete labels:

* **Seedlings**:  these data indicate spatial locations where seedlings of the tree have been found
* **Mortality**: these data indicate spatial locations where mortality has been observed
* **Persistence**: these data have no observed mortality or seedling at given spatial locations
* **Range**: the total dataset, including all spatial locations of seedlings, mortality, and persistence

---

## The Data

* The dataframe completed in exercise #6.

---

## Question #1

* Import and explore data
  * **NOTE**:  intent is not to reduce number of variables; that was completed in exercise #5.  Rather, calculate simple descriptive statistics (mean, sd, n) and boxplots


```{r}
# Load packages
library(tidyverse)
library(data.table)
library(PresenceAbsence) # PresenceAbsence for accuracy metrics
library(DAAG)
library(gam)             # smoothers with "s()" or LOWESS: "lo()"s
library(car)
library(raster)
library(sf)

# Set working directories
path.root <- "C:/Users/mttft/Desktop/UVM/UVM Courses/SDHM_R/sdhmR-V2021.1EXP"
path.mod2 <- paste(path.root, "/data/module02", sep = "")
path.figs <- paste(path.root, "/powerpoints/figures", sep = "") 
path.gis <- paste(path.root, "/data/gis_layers", sep = "")
path.range <- paste(path.root, "/data/RangeData", sep = "")

# Load CRSs
setwd(path.root)
source("r_code/prjs4_sdhmR.r")
ls(pattern = "prj.")

# Load presence:pseudo-absence and predictor data
load(paste(path.mod2, "/range_pres_psu_data.RData", sep = ""))
load(paste(path.mod2, "/tr_RANG.Rdata", sep = ""))
```

```{r}
tr_RANG_long <- tr_RANG %>% 
  pivot_longer(etpt_5:topos, names_to = "Variable", values_to = "Value") 
head(tr_RANG_long, 3)

#summarise data 
tr_RANG_long %>% group_by(Variable) %>% 
  summarise(n = n(),
            mean = mean(Value), 
            sd = sd(Value), 
            min = min(Value), 
            max = max(Value))

# see plots from ex 7
```

---

## Question #2

* Construct some basic plots to assess possible relationships warranting use of a GAM
* List which variables may be amenable to a GAM--based smoother

```{r}
# make scatter plots of variables (look for non-linearity)
plot(tr_RANG[18:24])

### make plots to identify response heterogeneity
## width of bins = data density
# par(mfcol = c(1, 3)) # if desired
col.list <- c("grey80", "forestgreen") # specified color ramp

plot(as.factor(tr_RANG$RANG106) ~ tr_RANG$etpt_5, xlab = "etpt_5", ylab = "Presence category", col = col.list) # plot 1: etpt_5

plot(as.factor(tr_RANG$RANG106) ~ tr_RANG$mind_yr_av, xlab = "mind_yr_av", ylab = "Presence category", col = col.list) # plot 2: mind_yr_av

plot(as.factor(tr_RANG$RANG106) ~ tr_RANG$prad_sw_di, xlab = "prad_sw_di", ylab = "Presence category", col = col.list) # plot 3: prad_sw_di

plot(as.factor(tr_RANG$RANG106) ~ tr_RANG$prec_w_hal, xlab = "prec_w_hal", ylab = "Presence category", col = col.list) # plot 4: prec_w_hal

plot(as.factor(tr_RANG$RANG106) ~ tr_RANG$rough_1k, xlab = "rough_1k", ylab = "Presence category", col = col.list) # plot 5: rough_1k

plot(as.factor(tr_RANG$RANG106) ~ tr_RANG$tmax_s_hal, xlab = "tmax_s_hal", ylab = "Presence category", col = col.list) # plot 6: tmax_s_hal

plot(as.factor(tr_RANG$RANG106) ~ tr_RANG$topos, xlab = "topos", ylab = "Presence category", col = col.list) # plot 7: topos
# par(mfrow = c(1, 1))

```
*Variables mind_yr_av, prec_w_hal, and tmax_s_hal would be most likely to need a GAM smoother. The remaining variables other than et_pt5 may all benefit from smoothers too*

---

## Question #3

* Construct a set of GAMs from fully linear to a maximum of 5 splines
  * Use lowess smoother rather than default spline

```{r}
# fully linear GAM
colnames(tr_RANG)
cn <- c(colnames(tr_RANG[,1:17]) ,paste(names(tr_RANG[,18:24]),".img", sep = ""))

colnames(tr_RANG) <- cn

PartRange_GAM <- gam(RANG106 ~ etpt_5.img+mind_yr_av.img+prad_sw_di.img+prec_w_hal.img+rough_1k.img+tmax_s_hal.img+topos.img, family = "binomial", data = tr_RANG) 
summary(PartRange_GAM)

# Mixed linear and smoothed model
PartRange_GAM2 <- gam(RANG106 ~ etpt_5.img+lo(mind_yr_av.img)+lo(prad_sw_di.img)+lo(prec_w_hal.img)+lo(rough_1k.img)+lo(tmax_s_hal.img)+lo(topos.img), family = "binomial", data = tr_RANG) 
summary.Gam(PartRange_GAM2)

# Stepwise GLM
PartRange_GAM3 <- step.Gam(PartRange_GAM,scope=list(
    "etpt_5.img" = ~1 + etpt_5.img + lo(etpt_5.img, 0.33) + lo(etpt_5.img, 0.2),
    "mind_yr_av.img" = ~1 + mind_yr_av.img + lo(mind_yr_av.img, 0.33) + lo(mind_yr_av.img, 0.2),
    "prad_sw_di.img" = ~1 + prad_sw_di.img + lo(prad_sw_di.img, 0.33) + lo(prad_sw_di.img, 0.2),
    "prec_w_hal.img" = ~1 + prec_w_hal.img + lo(prec_w_hal.img, 0.33) + lo(prec_w_hal.img, 0.2),
    "rough_1k.img" = ~1 + rough_1k.img + lo(rough_1k.img, 0.33) + lo(rough_1k.img, 0.2),
    "tmax_s_hal.img" = ~ 1 + tmax_s_hal.img + lo(tmax_s_hal.img, 0.33) + lo(tmax_s_hal.img, 0.2),
    "topos.img" = ~1 + topos.img + lo(topos.img, 0.33) + lo(topos.img, 0.2)), trace = F)
summary(PartRange_GAM3)
```

```{r}
PartRange_GAM2.fit <- 100 * (1 - PartRange_GAM2$deviance/PartRange_GAM2$null.deviance)
PartRange_GAM2.fit # 24.0 (up from full GLM)
PartRange_GAM2$deviance # 5,803.7 (down from full GLM)
PartRange_GAM2.pred <- predict(PartRange_GAM2, type = "response")
```

---

## Question #4

* Calculate accuracy metrics (as in Module 3.2 and 3.3, Analytical Intermission) using:
  * Resubstitution approaches, and
  * A 10-fold cross--validation approach

**Accuracy with Resubstitution**
```{r}
mod2 <- "GAM2"
dat2 <- cbind(mod2, tr_RANG[2], PartRange_GAM2.pred)
head(dat2, 2)

mod.cut <- optimal.thresholds(dat2, opt.methods = c("Default"))
mod.cut

# generate confusion matrix
PartRange_GAM2.cfmat <- table(dat2[[2]], factor(as.numeric(dat2$PartRange_GAM2.pred >= mod.cut$PartRange_GAM2.pred)))
PartRange_GAM2.cfmat

# calculate model accuracies with standard deviation=F
PartRange_GAM2.acc <- presence.absence.accuracy(dat2, threshold = mod.cut$PartRange_GAM2.pred, st.dev = F)
tss <- PartRange_GAM2.acc$sensitivity + PartRange_GAM2.acc$specificity - 1
PartRange_GAM2.acc <- cbind(PartRange_GAM2.acc[1:7], tss)
PartRange_GAM2.acc[c(1, 4:5, 7:8)]

auc.roc.plot(dat2, color = T)
```

**Accuracy with 10-fold Cross Validation**
```{r}
set.seed(1234)
PartRange_GAM2.cv <- CVbinary(PartRange_GAM2, nfolds = 10, print.details = F)
PartRange_GAM2.cv <- PartRange_GAM2.cv$cvhat
dat2 <- cbind(dat2[,1:2], PartRange_GAM2.cv)
head(dat2)

# generate confusion matrix
PartRange_GAM2.cfmatJ <- table(dat2[[2]], factor(as.numeric(dat2$PartRange_GAM2.cv >= mod.cut$PartRange_GAM2.pred)))
PartRange_GAM2.cfmatJ

# calculate model accuracy with standard deviation=F
PartRange_GAM2.accB <- presence.absence.accuracy(dat2, threshold = mod.cut$PartRange_GAM2.pred, st.dev = F)
tss <- PartRange_GAM2.accB$sensitivity + PartRange_GAM2.accB$specificity - 1
PartRange_GAM2.accB <- cbind(PartRange_GAM2.accB[1:7], tss)
PartRange_GAM2.accB[c(1, 4:5, 7:8)]

GAM2_acc.sum = rbind(PartRange.acc, PartRange.accB)
```


---

## Question #5

* Build 2 prediction maps:
  * A raw probability estimate for each cell in the modelling domain; and
  * A classified map based on the selected threshold from Question #4

```{r}
rang.probGAM <- predict(rangeDOM,PartRange_GAM2,filename="PartRange.probGAM.img",type="response",fun=predict,index=2, overwrite=TRUE)
rang.probGAM

setwd(path.gis)
states <- st_read(dsn = ".", layer = "na_states_wgs")

plot(rang.probGAM, legend = T, axes = T, main = "Probability Map")
plot(st_geometry(states), add = T, lwd = 1.5)
```

**Classified Map**
```{r classified map data}
rang.clasGAM <- reclassify(rang.probGAM, filename = "range.clas.GAM.img", 
   (c(0, mod.cut[[2]], 0, mod.cut[[2]], 1, 1)), overwrite=TRUE)

plot(rang.probGAM, legend = T, axes = T, main = "Classification Map")
plot(st_geometry(states), add = T, lwd = 1.5)
```



---

## Question #6

* Save your data as R objects:
  * Accuracy metrics as a dataframe;
  * Classification threshold as a stand--alone scalar
  * Both prediction maps as **`.img`** format
* Save these R objects in a **`.RData`** file

These data will be used again in Module 10, Ensemble Models.

```{r}
mod.cut_GAM <- mod.cut$PartRange_GAM2.pred

save(PartRange_GAM, PartRange_GAM2, GAM2_acc.sum, mod.cut_GAM, rang.probGAM, rang.clasGAM, file = "C:/Users/mttft/Desktop/UVM/UVM Courses/SDHM_R/sdhmR-V2021.1EXP/data/RangeData/GAM_Results.Rdata")
```

---

## The End

---
