---
title: "Exercise Title Goes Here"
author: "Your Name Goes Here"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: html_document
---

<!-- These are global options  
       set root directory here 
       NOTE mine is active; yours is commented out
       change the path to reflect your CPU,
       then comment out or delete mine
--> 
```{r global_options, include=FALSE}
knitr::opts_knit$set(root.dir = "~/words/classes/sdmR_ALLversions/sdhmR-V2020.2")
#knitr::opts_knit$set(root.dir = "~/sdhmR-V2020.2")
knitr::opts_chunk$set(warning=FALSE,error=TRUE,message=FALSE)
```

---

## This exercise links to Module 6

Submit the completed exercise to us at:

* t.edwards@usu.edu 
* eric_tsakiris@fws.gov  

---

## Context

The data are of the *Pinus edulis* Common piñon, the species you dealt with in Exercise #3 -- #5.  This exercise relies on the dataframe you generated in exercise #6.  Hopefully you saved those combined presence:absence dataframe, with the reduced variables you begin SDHM model construction with, or regrettably you will need to return to exercise #6 and re--build that dataframe.

The goal is to: 

* Build ensemble map products of the $\bar{x}$ and $SD$ probabilities 
* Build ensemble concordance map products of 3, 5, and union of all models 
* Build tables of frequencies: 
  * Presences (ie, sum the cells of classified presence for the 3, 5, and union ensembles) 
  * Concordances (remember, this is a measure of "how much !!) 

Remember, you will have been assigned data for *edulis* having one the following four discrete labels:

* **Seedlings**:  these data indicate spatial locations where seedlings of the tree have been found
* **Mortality**: these data indicate spatial locations where mortality has been observed
* **Persistence**: these data have no observed mortality or seedling at given spatial locations
* **Range**: the total dataset, including all spatial locations of seedlings, mortality, and persistence

---

## The Data

* Load:
  * The previously constructed SDHM models, for your group label, and 
  * The (i) prediction; and (ii) classified maps

If you did **NOT** save these you're in a bit of trouble ...

---

## Question #1

* Build ensemble map products of:
  * SDHM mean and sd probabilities
  * SDHM concordance maps of 5, 3, and union of all SDHM overlaps
  * "Clip" all these maps by the bounding boxes created earlier (see Module 2.3.3 for refresher, if needed)
  * Output these map products as **`.img`** files

```{r}
# code for the question goes inside chunks, if part of the exercise solution.
#  If not part of the exercise questions delete this.
#  NOTE: do NOT answer all questions from within the same chunk
```

**YET ANOTHER NOTE RELATED TO CODE:**  
If King's English interpretation is required, it will go here, after each code chunk.

---

## Question #2

* Calculate the frequencies of "presence" in each of the 5 SDHMs

---

## Question #3

* Tally the frequencies of concordance after "clipping" and compare with above

---

## Question #4

* Save your data as R objects:
  * All ensemble prediction maps as **`.img`** format
* Save these R objects in a **`.RData`** file as well

---

## The End

---
