---
title: "Exercise 2"
author: "Eli Polzer, Mia McReynolds, Matt Futia"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  pdf_document: default
  html_document: default
---


```{r global_options, include=FALSE}

```

---

## This exercise links to Module 2.3.3

Submit the completed exercise to us at:

* t.edwards@usu.edu 
* eric_tsakiris@fws.gov 

---

## Context

The exercise explores the consequences of frame selection on numerical summaries of predictive variables.  The species being explored is [*Pinus edulis* Common piñon](https://www.conifers.org/pi/Pinus_edulis.php).  Feel free to refer to the illustrations in Module 2.3.2 for graphical context.

## The Data

* Load **`spp106.soil.Rdata`** and **`spp106.clim.Rdata`** (or import **`spp106.soil.csv`**, **`spp106.clim.csv`**)
* Data in: **`~\sdhmR-V2020.2\data\module02`**

```{r}
load("C:/Users/mttft/Desktop/UVM/UVM Courses/SDHM_R/sdhmR-V2021.1EXP/data/module02/spp106.soil.Rdata")
load("C:/Users/mttft/Desktop/UVM/UVM Courses/SDHM_R/sdhmR-V2021.1EXP/data/module02/spp106.clim.Rdata")
library(tidyverse)
```


---

## Question #1

* What are the expected consequences in terms of a SDHM classification model if differences between environmental variables measured at presence:absence sites are large vs. small?

MHF: *Environmental variables with large differences would require a low resolution that may not distinguish factors with small variance. As a result, these factors with smaller variance, which may have high biological importance, would have less influence for modeling a species distribution.*

---

## Question #2

* How might these differing frames, and their respective variable differences based on the presence or absence of the species, affect: 
  * (i) ecological interpretations, and 
  * (ii) their classification ability, of variables?

MHF: *i: Having frames that are much larger than the actual range of a species may cause ecological interpretations to be biased towards environmental factors with large variance rather than factors that are most influential for the habitat use. Alternatively, small frames may exclude observations resulting in a different interpretation of the species' habitat use.*
*ii: Having biases favoring environmental factors with greater variance may reduce classification ability by minimizing the influence of factors that are most relevant for describing a species' habitat.*

---

## Question #3

* How do you decide which frame -- given many competing -- is appropriate for SDHM construction?

MHF: *Based on the frame that incorporates the greatest number of environmental variables associated with a species of interest, while also trying to incorporate the maximum geographic space used by the species.*

---

## Question #4

* Examine variable names and select 3 for all subsequent questions.  Identify them below in your answer to this question.

```{r}
colnames(spp106.clim)
colnames(spp106.soil)
```

* *awc_wtavg*
* *clay*
* *frag3to10*

---

## Question #5

* Run some simple descriptive statistics (ie, $\bar{x}$, $SD$) on the three (3) selected variables from above and compare across frames (where 1=inside frame; 0=outside): 
* There are four (4) frames to evaluate:
  * Western U.S. states; this frame is defined as the entire dataframe N
  * **`$dry.domfr`** = all *edulis* within the dry domain
  * **`$SPPRES106.fiafr`** = buffered *edulis* distribution based on FIA data points
  * **`$SPPRES106.cmbfr`** = buffered *edulis* distribution based on FIA data points plus the UNION with [Little’s mapped tree distributions](https://www.fs.fed.us/database/feis/pdfs/Little/aa_SupportingFiles/LittleMaps.html).
  
### *Western US frame:*
```{r}
all = spp106.soil[,c("awc_wtave", "clay", "frag3to10")]
apply(all,2,function(all){mean(all, na.rm = T)})
apply(all,2,function(all){sd(all, na.rm = T)})
```
* *awc_wtave: 2843.6 $\pm$ 2412.3*
* *clay: 21.2 $\pm$ 9.4*
* *frag3to10: 6.5 $\pm$ 8.1*

### *`$dry.domfr` frame:*
```{r}
dry = filter(spp106.soil[,c("dry.domfr","awc_wtave", "clay", "frag3to10")], dry.domfr == 1)
apply(dry,2,function(dry){mean(dry, na.rm = T)})
apply(dry,2,function(dry){sd(dry, na.rm = T)})
```
* *awc_wtave: 2915.0 $\pm$ 2443.0*
* *clay: 21.2 $\pm$ 9.4*
* *frag3to10: 6.6 $\pm$ 8.2*

### *`$SPPRES106.fiafr` frame:*
```{r}
fiafr = filter(spp106.soil[,c("SPPRES106.fiafr","awc_wtave", "clay", "frag3to10")], SPPRES106.fiafr == 1)
apply(fiafr,2,function(fiafr){mean(fiafr, na.rm = T)})
apply(fiafr,2,function(fiafr){sd(fiafr, na.rm = T)})
```
* *awc_wtave: 3356.0 $\pm$ 2564.7*
* *clay: 21.0 $\pm$ 7.2*
* *frag3to10: 8.7 $\pm$ 9.8*

### *`$SPPRES106.cmbfr` frame:*
```{r}
cmbfr = filter(spp106.soil[,c("SPPRES106.cmbfr","awc_wtave", "clay", "frag3to10")], SPPRES106.cmbfr == 1)
apply(cmbfr,2,function(cmbfr){mean(cmbfr, na.rm = T)})
apply(cmbfr,2,function(cmbfr){sd(cmbfr, na.rm = T)})
```
* *awc_wtave: 3353.5 $\pm$ 2572.6*
* *clay: 20.8 $\pm$ 7.2*
* *frag3to10: 8.5 $\pm$ 9.7*

---
  
## Question #6

* Build box plots with the 4 frames by variable for each of your selected variable, ie this will be 3 variables $\times$ 4 different frames.

```{r}
par(mfrow = c(2,2))
boxplot(all)
boxplot(dry[2:4])
boxplot(fiafr[2:4])
boxplot(cmbfr[2:4])
```


---

## Question #7: 

* Identify which of the variables that subjectively (No statistical tests here, please!Just your Mark I eyeballs.) look as though they would discriminate between presence and absence

MHF: *The awc_wtave looks like it would have the most influence between presence and absence because it has the greatest variance for all frames*

---

## The End

---