---
title: "Homework 3: Modelling Frame and FISHNET Construction"
author: "Eli Polzer, Matt Futia, and Amelia McReynolds"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: html_document
---

```{r global_options, include=FALSE}
#Mia's directory
knitr::opts_knit$set(root.dir = "~/University of Vermont/Classes/sdhmR/sdhmR-V2021.1")

knitr::opts_chunk$set(warning=FALSE,error=TRUE,message=FALSE)

#filepaths
  path.root <- "~/University of Vermont/Classes/sdhmR/sdhmR-V2021.1" 
  path.mod2 <- paste(path.root, "/data/module02", sep = "")
  path.figs <- paste(path.root, "/powerpoints/figures", sep = "") 
  path.gis <- paste(path.root, "/data/gis_layers", sep = "")
```

```{r}
# some libraries
  library(raster)	# FXNS: extent, raster, res, rasterize, values, writeRaster, 
					#       cellFromXY, extract
  library(sf)    	# FXNS: st_as_sf, st_as_sfc, st_bbox, st_convex_hull, st_union
					#       st_buffer, st_read, st_write, st_intersection, st_geometry,
					#       st_drop_geometry, st_coordinates, st_sf
  library(terra)	# FXNS: as.data.frame, rast, vect
  library(readr)
  library(ggplot2)
  library(dplyr)
  library(tidyr)
```


---

## This exercise links to Module 2.3.4

Submit the completed exercise to us at:

* t.edwards@usu.edu 
* eric_tsakiris@fws.gov 

---

## Context

The Uinta Basin hookless cactus (*Sclerocactus wetlandicus*) (SCWE) is listed as a threatened species.  Recovery goals require:
* An accurate range map;
* Comprehensive surveys on locations; and
* Other needs as well
[FWS ECOS](https://ecos.fws.gov/ecp/species/9037) has an expert opinion-based "blob" of an hypothetical SCWE modelling (ie, the "blob") domain.

## The Data

* Acquire from the ECOS site the SCWE "blob" shapefile (call it: scwe_ecospoly) 
* Import the **`scwe_pres.csv`** file of known SCWE presence points
* Data in: **`~\sdhmR-V2020.2\data\gis_layers`** and **`...\module02`**, respectively

---

## Question #1

* Import the ECOS--acquired SCWE "blob" as a spatial object (call it: **`scwe_ecospoly`**)
* Find and acquire the **`CRS epsg:42303`** and it's **`proj4string`** – this is the CRS for ALL subsequent analyses
* Set CRS of (i) the ECOS "blob" and (ii) the imported points to epsg:42303
* Plot the presence points relative to the ECOS-based polygon "blob"

```{r}
#import ECOS blob shapefile w/ sf package
scwe_ecospoly <- st_read("~/University of Vermont/Classes/sdhmR/sdhmR-V2021.1/data/gis_layers/scwe_ecospoly.shp")

#import presence points data w/ readr
scwe_pres <- read_csv("~/University of Vermont/Classes/sdhmR/sdhmR-V2021.1/data/module02/scwe_pres.csv", 
     col_types = cols(date.r = col_date(format = "%Y-%m-%d")))
```

```{r}
#proj4string for epsg:42303 from https://epsg.io/42303 
prj.epsg42303 <- "+proj=aea +lat_1=29.5 +lat_2=45.5 +lat_0=23 +lon_0=-96 +x_0=0 +y_0=0 +ellps=GRS80 +datum=NAD83 +units=m +no_defs no_defs"

#proj4string for the scwe_pres current CRS:  ALBERS NORTH AMERICA NAD83
#   https://epsg.io/5070
    prj.aeaN83 <- "+proj=aea +lat_1=29.5 +lat_2=45.5 +lat_0=23 +lon_0=-96 +x_0=0 +y_0=0 +ellps=GRS80 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs" # epsg:42303 


#set CRS of the ECOS "blob" to epsg:42303 -- convert from the current crs
  scwe_ecospoly_01 <- st_transform(scwe_ecospoly, crs = prj.epsg42303) # output pts as spatial
  st_crs(scwe_ecospoly_01)$proj4string # check crs 

  
#set CRS of the imported points to epsg:42303 - similar to CSV in m02.01-dataimport
pts.in <- st_as_sf(scwe_pres, coords = c("aeaN83_x", "aeaN83_y"), 
    crs = prj.aeaN83, remove = F) # remove=F to retain input x,y
  st_crs(pts.in)$proj4string # check crs

  # transform to new crs -- convert from the current crs
  scwe_pres_01 <- st_transform(pts.in, crs = prj.epsg42303) # output pts as spatial
  st_crs(scwe_pres_01)$proj4string # check crs -looks good

  # adding new geometry back into dataframe as columns
  coords <- data.frame(st_coordinates(scwe_pres_01)) # create df of new geometry
  names(coords) <- c("NAD83_x", "NAD83_y") # assign names
  scwe_pres_01 <- st_sf(data.frame(scwe_pres_01,coords)) # add columns

```

```{r}
# Plot the presence points relative to the ECOS-based polygon "blob"
ggplot() +
  geom_sf(data=scwe_ecospoly_01, fill="gray70") +
  geom_point(data=scwe_pres_01, aes(x=aeaN83_x, y=aeaN83_y), color="darkgreen") 
```

---

## Question #2

* Generate a buffered point-based (ie OPTION #4 from Mod 2.3.4) modelling frame
* Set buffer of 10,000 m
* Set CRS of modelling frame to epsg:42303

```{r}
# obtain desired crs for modelling frame & res from existing GIS data layer in NAD83 (epsg42303)
  t1 <- raster("~/University of Vermont/Classes/sdhmR/sdhmR-V2021.1/data/gis_layers/elev_1k_aea.img")
crs(t1)
```

```{r}
#make a buffered sf 
scwe.bufptSF <- st_buffer(scwe_pres_01, dist = 10000) # 10,000 m buffer (units are meters)
  scwe.bufptSF <- st_union(scwe.bufptSF)  # merges all intersecting polys - usual for SDHM
  scwe.bufptSF # examine
  
# extract buffered points from template & rasterize
  scwe.bufptSP <- as_Spatial(scwe.bufptSF) # convert to sp spatialpolygon class
  scwe.bufptR <- crop(rasterize(scwe.bufptSP, t1), scwe.bufptSP) # t1 is template from above
  scwe.bufptR # examine
```

---

## Question #3

* Select larger of the two frames as FISHNET extent 
  * (HINT: examine extents of both -- one will be larger)
  
```{r}
  ggplot() +
  geom_sf(data=scwe_ecospoly_01,  fill="gray70") +
    geom_sf(data=scwe.bufptSF, color="black", lwd=1.5, fill=NA)+
  geom_point(data=scwe_pres_01, aes(x=NAD83_x, y=NAD83_y), color="darkgreen") +
  theme_classic()
    
```
  The frame created with 10 km-buffered points is larger than the ECOS polygon, so I will use the buffered points shapefile as the FISHNET extent. 

---

## Question #4

* Build a 1,000 m (1 km) square FISHNET
* Save both spatial and dataframe FISHNETs as R objects 
  * (NOTE: we will use these later so be sure and save !!!)
* Export as both dataframe (ie, no geometry) and ESRI shapefile


```{r}

# extract coords, cell id of coords, & build dataframe
  f1 <- sp::coordinates(scwe.bufptR) # get spatial coords from scwe.bufptR
  f2 <- cellFromXY(scwe.bufptR, f1) # extract cell number from buffered extent
  f3 <- as.data.frame(cbind(f1, f2)) # build datframe of x,y & cell number
  head(f3, 2) # examine
  tail(f3, 1) # f2 value should eq. ncell in scwe.bufptR; below
  ncell(scwe.bufptR) # should eq. f2 value above
# some dataframe clean-up
  names(f3)[1:3] <- c("cell.wgs_x", "cell.wgs_y", "FNETID") # assign names
  f3 <- f3[c("FNETID", "cell.wgs_x", "cell.wgs_y")] # reorder

# examine fishnet data structure
  dim(f3) # dimension; rows is maximum number of cells in fishnet 
  names(f3) # names in fishnet
  head(f3, 2) # note first FNETID number
  tail(f3, 2) # note last FNETID number matches dimension size

# convert dataframe to spatial object
  scwe.fnetSF <- st_as_sf(f3, coords = c("cell.wgs_x", "cell.wgs_y"), 
    crs = prj.epsg42303, remove = F) # remove=F retains input x,y)
  head(scwe.fnetSF, 2) # examine

# build stand-alone dataframe from geospatial object
  class(scwe.fnetSF) # note dataframe class; will extract data.frame
  scwe.fnetDF <- st_drop_geometry(scwe.fnetSF) # build dataframe
  head(scwe.fnetDF, 2) # examine
  class(scwe.fnetDF) # check if class=data.frame

# export as esri shapefile & dataframe
  setwd(path.mod2) # output path
  st_write(scwe.fnetSF, dsn = ".", layer = "scwe_fnetSF", driver = "ESRI Shapefile",
    delete_layer = T, delete_dsn = T) # output shapefile

# save objects
  setwd(path.mod2) # output path
  save("scwe.fnetSF", "scwe.fnetDF", file = "scwe.fnet.RData")
```


---

## Question #5

Once you have completed the analyses above, answer the following:

* How do the known SCWE presences relate to the modelling frame "blob"?
* How different is the SCWE "blob" from a buffered point-based modelling domain?
* What is the best "blob" for creating a data indexing FISHNET?  Why?


The known presences were used to create our modelling frame blob. We used a 10 km buffer around the points to make a FISHNET. 

The ECOS polygon was similar to the buffered point-based modelling domain that we used, but was smaller and the boundaries were more irregular. Maybe it was traced by hand in ArcGIS?

The best blob matches the objectives set out in your Intended Use Statement, and uses resolution that matches the resolution of your data somewhat closely. Sweeping points into a FISHNET obscures their specific location and reports them as a value assigned to a cell's centroid, so it's important to avoid losing too much resolution from your data. 

---

## The End

---
