---
title: "Exercise Title Goes Here"
author: "Your Name Goes Here"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: html_document
---

<!-- These are global options  
       set root directory here 
       NOTE mine is active; yours is commented out
       change the path to reflect your CPU,
       then comment out or delete mine
--> 
```{r global_options, include=FALSE}
knitr::opts_knit$set(root.dir = "~/words/classes/sdmR_ALLversions/sdhmR-V2020.2")
#knitr::opts_knit$set(root.dir = "~/sdhmR-V2020.2")
knitr::opts_chunk$set(warning=FALSE,error=TRUE,message=FALSE)
```

---

## This exercise links to Module 2.3.4

Submit the completed exercise to us at:

* t.edwards@usu.edu 
* eric_tsakiris@fws.gov 

---

## Context

The Uinta Basin hookless cactus (*Sclerocactus wetlandicus*) (SCWE) is listed as a threatened species.  Recovery goals require:
* An accurate range map;
* Comprehensive surveys on locations; and
* Other needs as well
[FWS ECOS](https://ecos.fws.gov/ecp/species/9037) has an expert opinion-based "blob" of an hypothetical SCWE modelling (ie, the "blob") domain.

## The Data

* Acquire from the ECOS site the SCWE "blob" shapefile (call it: scwe_ecospoly) 
* Import the **`scwe_pres.csv`** file of known SCWE presence points
* Data in: **`~\sdhmR-V2020.2\data\gis_layers`** and **`...\module02`**, respectively

---

## Question #1

* Import the ECOS--acquired SCWE "blob" as a spatial object (call it: **`scwe_ecospoly`**)
* Find and acquire the **`CRS epsg:42303`** and it's **`proj4string`** – this is the CRS for ALL subsequent analyses
* Set CRS of (i) the ECOS "blob" and (ii) the imported points to epsg:42303
* Plot the presence points relative to the ECOS-based polygon "blob"

```{r}
# code for the question goes inside chunks, if part of the exercise solution.
#  If not part of the exercise questions delete this.
#  NOTE: do NOT answer all questions from within the same chunk
```

**YET ANOTHER NOTE RELATED TO CODE:**  
If King's English interpretation is required, it will go here, after each code chunk.

---

## Question #2

* Generate a buffered point-based (ie OPTION #4 from Mod 2.3.4) modelling frame
* Set buffer of 10,000 m
* Set CRS of modelling frame to epsg:42303

---

## Question #3

* Select larger of the two frames as FISHNET extent 
  * (HINT: examine extents of both -- one will be larger)

---

## Question #4

* Build a 1,000 m (1 km) square FISHNET
* Save both spatial and dataframe FISHNETs as R objects 
  * (NOTE: we will use these later so be sure and save !!!)
* Export as both dataframe (ie, no geometry) and ESRI shapefile

---

## Question #5

Once you have completed the analyses above, answer the following:

* How do the known SCWE presences relate to the modelling frame "blob"?
* How different is the SCWE "blob" from a buffered point-based modelling domain?
* What is the best "blob" for creating a data indexing FISHNET?  Why?

---

## The End

---
