---
title: "Exercise 5: Building the Training Dataframe"
author: "Group A: Eli Polzer, Matt Futia, Mia McReynolds"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: html_document
---

```{r global_options, include=FALSE}
knitr::opts_knit$set(root.dir = "~/University of Vermont/Classes/sdhmR/sdhmR-V2021.1")
knitr::opts_chunk$set(warning=FALSE,error=TRUE,message=FALSE)
```

```{r}
# some pathnames; yours will be specific to your CPU !!
  path.root <- "~/University of Vermont/Classes/sdhmR/sdhmR-V2021.1" 
  
# below is the recommended class root dir
  #path.root <- "~/sdhmR-V2021.1"  # typical class root dir
  path.mod2 <- paste(path.root, "/data/module02", sep = "")
  path.figs <- paste(path.root, "/powerpoints/figures", sep = "") 
  path.gis <- paste(path.root, "/data/gis_layers", sep = "")

# some libraries
  library(raster)	# FXNS:	raster, projection, projectRaster, extract, crop
  library(readr)
  library(ggplot2)
  library(dplyr)
   library(sf)    
```


```{r, message=FALSE}
#import range-wide data
  spp106pr_RANG <- read_csv("~/University of Vermont/Classes/sdhmR/sdhmR-V2021.1/data/exercise/traindat/spp106pr_RANG.csv")

#add proj string for WGS 84
prj.wgs84 <- "+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs" # epsg:4326

  
#load presence/absence data 
  # setwd(path.mod2) - not working?
  pres.abs <- get(load("~/University of Vermont/Classes/sdhmR/sdhmR-V2021.1/data/module02/pied.PPsA.RData")) 
  # load presence:absence data
  head(pres.abs, 2) # examine
```
  
```{r}
  load("~/University of Vermont/Classes/sdhmR/sdhmR-V2021.1/data/module02/pied.framesR.RData") 

  pied.bufptR # examine cropping raster
  plot(pied.bufptR) # examine raster plot
```


## The Data

* Common piñon *Pinus edulis* sppcode=106
* The pseudo-absence data: Wherever stored from Homework #4
* Predictors: **`~\sdhmR-V2020.2\data\exercise\preds`** 
  * There are a total of 16 predictor variables the directory above
  * The variables are **`.img`** files
* Data files:
  * **`spp106pr_MORT.csv`**, **`spp106pr_PERS.csv`**, **`spp106pr_RANG.csv`**, **`spp106pr_SEED.csv`**

---

## Question #1

* Import the predictor GIS variables into R

```{r, message=FALSE}
#run this entire chunk at one time
 setwd("~/University of Vermont/Classes/sdhmR/sdhmR-V2021.1/data/exercise/preds")
img.list <- list.files(pattern = ".img$") # list of .img files; $ strips extra
img.list # examine

# loop for extracting topo data w/raster stack
  layers <- {} # initialize (empty) list of raster layers
  for (i in 1:length(img.list)) {
    r1 <- crop(raster(img.list[i]), pied.bufptR) # crop pred var raster to buffer raster
    names(r1) <- strsplit(img.list[i], "_wgs.img") # assign name to raster
   layers <- c(layers, r1) # add raster to layer of rasters
  }
  layers # examine raster stack; return is a list of raster layers
```

It looks like all CRS's of rasters are WGS84, so I won't have to re-project. Also all resolutions are the same, 0.0083 x 0.0083. 

---

## Question #2

* Extract predictor variables from all 16 **`.img`** files using the presence:absence [X,Y]'s

```{r}
# build the raster stack
  pied.topoDOM <- stack(layers) # create a raster stack
  pied.topoDOM # examine stack

# extract from stack 
  t1 <- extract(pied.topoDOM, pres.abs[, c("tr.wgs_x", "tr.wgs_y")]) # extract values from raster stack
  head(t1, 2) # examine extracted matrix
```
Looks good: all 16 rasters are there, and resolution/CRS are preserved. 

---

## Question #3

* Create a new dataframe by binding the extracted predictor variable values to the dataframe from exercise #4.

```{r}
  pied.trTOPO <- cbind(pres.abs, t1) %>% # bind to train dataframe
    select(-c(wgs_xF, wgs_yF, cell.wgs_x, cell.wgs_y))

  head(pied.trTOPO, 2) # examine training data frame
```

I removed the wgs_xF and cell.wgs_x columns, since all coords are in tr.wgs_x. 

```{r}
ggplot(pied.trTOPO, aes(x=tr.wgs_x, y=tr.wgs_y)) +
  geom_point(aes(color=prec_winte.img)) +
  theme_classic()
```

A plot for fun, winter precipitation. Looks reasonable. 

```{r}
ggplot(pied.trTOPO, aes(x=tr.wgs_x, y=tr.wgs_y)) +
  geom_point(aes(color=rough_1k.img)) +
  theme_classic()
```

Assuming this is some kind of elevation layer? Looks like some of these predictors will be correlated. 

```{r}
ggplot(pied.trTOPO, aes(x=tr.wgs_x, y=tr.wgs_y)) +
  geom_point(aes(color=factor(SPPRES106))) +
  theme_classic()
```
Based on these maps, this species prefers higher-elevation sites. 

---

## Question #4

* Save your data as R objects:
  * Dataframe;
  * Point shapefile with geometry in R
  * Export as a point shapefile in ESRI format
* Save these R objects in a **`.RData`** file

These data will be used in the next exercise #6.

```{r}

# write out data files if desired
  setwd(path.mod2)
  write.csv(pied.trTOPO, file = "pied_trTOPO.csv", row.names = F) # save .csv - training data
  save(pied.trTOPO, file = "pied.trTOPO.RData") # save .RData - pres/abs plus stacked raster
  save(pied.topoDOM, file = "pied.topoDOM.RData") # save .RData - stacked raster
  st_write(pied.trTOPO, dsn = ".", layer = "pied.trTOPO", driver = "ESRI Shapefile",
    delete_layer = T, delete_dsn = T)#point shapefile in ESRI format
```

---

## The End

---
